taxize - taxonomic search and retrieval in R

### 2013-05-28 inst/doc/taxize_withcode.Rnw
added bibliographystyle to fix issue #15
typos

\author{Eduard Sz\"{o}cs}
\email[E-mail: ]{szoe8822@uni-landau.de}
\affiliation{Institute for Environmental Sciences, University Koblenz-Landau, Forststraße 7, 76829 Landau, Germany}
\keywords{taxonomy; R; software; data; API}

@@

Evolution by natural selection has led to a hierarchical relationship among all living organisms.  Thus, species are categorized using a taxonomic hierarchy, starting with the binomial species name (e.g, \emph{Homo sapiens}), moving up to genus (\emph{Homo}), then family (\emph{Hominidae}), and on up to Domain (\emph{Eukarya}). Biologists, whether studying organisms at the cell, organismal, or community level, can put their study taxa into taxonomic context, allowing them to know close and distant relatives, find relevant literature, and more. Discovering the correct taxonomic names is, unfortunately, not straightforward. Taxonomic names often change due to name changes at the generic or specific levels, lumping or splitting lower taxa (genera, species) among higher taxa (families), and name spelling changes. In addition, there is no one authoritative taxonomic names source. Instead, there are essentially competing sources (e.g., uBio, Tropicos, ITIS) that may have different accepted names for the same taxon. The goal of taxize, an R package in development, is to make all use cases having to do with retrieving and resolving taxonmic names easy and replicable.

@@

One data source available in taxize is available in another R pacakge that solely interact with that data source. We provide a few convenience functions that wrap functions in \emph{taxonstand}, an R interface to The Plant List \cite{theplantlist}.

@@

taxize takes the approach that the user should be able to make decisions about what resource to trust, rather than taxize making the decision. Both the EOL GNR and the TNRS services provide data from a variety of data sources. The user may trust a specific data source, thus may want to use the names from that data source. In the future, we may provide the ability for taxize to suggest the best match from a variety of sources, but since R is relatively inefficient in memmory management, etc., we would rather offload this sort of computationally intensive task.

@@

It turns out that the panther has a statu of endangered (EN) and the lynx has a status of least concern (LN). (ED, WHAT ELSE DO YOU WANT TO SAY HERE?)

@@

Another use case available in taxize deals with genetic sequences. taxize has three functions to interact with GenBank to search for available genes (\emph{get\_genes\_avail}), download genes by GenBank ID (\emph{get\_genes}), and download genes via taxonomic name search, including retrieving a congeneric if the searched taxon does not exist in the databas (\emph{get\_seqs}). In this example, we search for gene sequences for \emph{Umbra limi}.

@@

The taxize package is part of the rOpenSci project \url{http://ropensci.org/}. We thank X, Y, and Z for comments on previous versions of this manuscript. SAC is supported by CANPOLIN of Canada, grant number XXXXXX.

>>>>

\author{Eduard Sz\"{o}cs}
\email[E-mail: ]{szoe8822@uni-landau.de}
\affiliation{Institute for Environmental Sciences, University Koblenz-Landau, Fortstraße 7, 76829 Landau, Germany}
\keywords{taxonomy; R; software; data; API}

@@

Evolution by natural selection has led to a hierarchical relationship among all living organisms.  Thus, species are categorized using a taxonomic hierarchy, starting with the binomial species name (e.g, \emph{Homo sapiens}), moving up to genus (\emph{Homo}), then family (\emph{Hominidae}), and on up to Domain (\emph{Eukarya}). Biologists, whether studying organisms at the cell, organismal, or community level, can put their study taxa into taxonomic context, allowing them to know close and distant relatives, find relevant literature, and more. Discovering the correct taxonomic names is, unfortunately, not straightforward. Taxonomic names often change due to name changes at the generic or specific levels, lumping or splitting lower taxa (genera, species) among higher taxa (families), and name spelling changes. In addition, there is no one authoritative taxonomic names source. Instead, there are essentially competing sources (e.g., uBio, Tropicos, ITIS) that may have different accepted names for the same taxon. The goal of taxize, an R package in development, is to make all use cases having to do with retrieving and resolving taxonomic names easy and replicable.

@@

One data source available in taxize is available in another R package that solely interact with that data source. We provide a few convenience functions that wrap functions in \emph{taxonstand}, an R interface to The Plant List \cite{theplantlist}.

@@

taxize takes the approach that the user should be able to make decisions about what resource to trust, rather than taxize making the decision. Both the EOL GNR and the TNRS services provide data from a variety of data sources. The user may trust a specific data source, thus may want to use the names from that data source. In the future, we may provide the ability for taxize to suggest the best match from a variety of sources, but since R is relatively inefficient in memory management, etc., we would rather offload this sort of computationally intensive task.

@@

It turns out that the panther has a status of endangered (EN) and the lynx has a status of least concern (LN). (ED, WHAT ELSE DO YOU WANT TO SAY HERE?)

@@

Another use case available in taxize deals with genetic sequences. taxize has three functions to interact with GenBank to search for available genes (\emph{get\_genes\_avail}), download genes by GenBank ID (\emph{get\_genes}), and download genes via taxonomic name search, including retrieving a congeneric if the searched taxon does not exist in the database (\emph{get\_seqs}). In this example, we search for gene sequences for \emph{Umbra limi}.

@@

The taxize package is part of the rOpenSci project \url{http://ropensci.org/}. We thank X, Y, and Z for comments on previous versions of this manuscript. SAC is supported by CANPOLIN of Canada, grant number XXXXXX. We thank all API maintainers for their work making their databases open to the public.
